The effects of symbols and information in government communication on citizens  trust in policy: Evidence from a survey experiment

### 2018-05-21 paper.Rmd
Add files via upload

Preliminary draft. Last edited at `r Sys.Date()`

@@

I conducted the survey experiment through Qualtrics Survey Software. To select and recruit residents of these specific cities, I was assisted by an Israeli internet research panel company *iPanel*. The online survey link was sent by iPanel, between 28.1-1.2.2018. A total of `r N_0` respondents completed the survey (a response rate of `r round((N_0/5077)*100,1)`%).[^footnote_responserate] Of these, I later filtered out `r N_0-N_1` observations (`r round(((N_0-N_1)/N_0)*100,1)`%) due to multiple entries from the same IP address (n=`r filter_IP`), failure at the instructional manipulation test (n=`r filter_IMC`), submitting the survey in less than 3 or more than 30 minutes (n=`r filter_time`), and respondents under the age of 18 (n=`r filter_age`). These filters are not significantly associated with the areas or the experimental conditions. Ultimately, following this rigorous screening, I ended up with a sizable sample of `r N_1` subjects, of whom `r filter(SVIVA2_01,AREA_reside_haifa==1)%>%nrow()` reside in the Haifa-bay cities, and `r filter(SVIVA2_01,AREA_reside_haifa==0)%>%nrow()` reside in the center cities.

@@

<br>
**Figure 3: Experimental design for policy plan communications**
![](D:/UNIVERSITY/R/SVIVA/experiment 2 01-2018/myfigures/experiment_flow.png)

@@

A growing body of Public-Administration research seeks to assess whether and how government organizations can increase citizens' trust through direct communications with citizens. This study contributes to this discussion, by pointing to the influence of symbolic elements in public communications, and unpacking the theoretical mechanisms underlying their effect on citizens' trust. Using a rigorous experimental design, embedded in a realistic setting of the Israeli Environmental Protection Ministry's communication of policy plans, I tested the effect of symbolic elements on trust, and its interactions with the substantive content of the communications and with their perceived personal relevance. To the best of my knowledge, this is the first study to test these interactions in the context of government communications.

The experimental results demonstrate that using familiar symbolic elements in communications about policy plans increase citizens' trust in these policies, via two supplementary mechanisms: First, they evoke general positive feelings that are transferred to the specific policy plan. Second, by putting citizens in a positive mood, they distract them from thinking about the substantive content of the communication, and thereby, they cause them to disregard its weaknesses. The second mechanism entails an interaction between symbols and weak information. These two mechanisms can be understood in light of the social psychology elaboration likelihood model. I theorized, based on ELM, that the effect of symbols is attenuated the more the communication is personally relevant for citizens' lives, but this expectation is not supported by the results. Rather, I found that citizens who perceived the policy issue as highly relevant for their lives (i.e. air-pollution in their area of residence), were similarly distracted by the symbols (and other peripheral cues), and persuaded to trust government's actions, even when they were undoubtedly illogical. These findings are based on a large sample of more than `r (SVIVA2_01%>%nrow() %>% round(-2))-100` participants, and are robust for various model specifications.

How can we explain the null findings about personal relevance?  
- very low motivation, even among Haifa-Bay subjects. Not new, existing problem. congruent with IEC study.

Limitation of study:
- one case...
- personal relevance based on observational.

What are the implications of these findings for public administration research? First, they highlight the important role of symbols and brands in shaping citizens' attitudes about government. Citizens' responses to symbolic elements in communications thus far received relatively little attention in Public administration literature, compared with the the abounded research on responses to information. Yet, this study, and others, demonstrated that symbols can be fairly effective in increasing citizens' trust, and mitigating their negative attitudes [@alonbarkat_2017; @karens_2016; @teodoro_2018]. Furthermore, this study shows that symbols decrease the "rational", systematic processing of substantive information. This may suggest that government organizations bypass or at least attenuate citizens' negative biases that undermine the effectiveness of positive information, by employing symbolic communication.[^footnote_biases_ELM] This is good news as far as we are concerned mainly about citizens' *undervaluation* of government performance. On the other hand, even if we assume that public organizations provide excellent outcomes, the emotive effect of government symbols on public opinion is still problematic, insofar as it challenges the democratic ideal that citizens form their opinions about their government and its actions on the basis of rational thinking [@dahl_1973; @soroka_2010]. Moreover, it becomes especially problematic when symbols are used by poorly performing organizations,  or organizations that lack professional expertiese. It those cases, symbolic elements can mitigate justified public criticism of government failures. Organizations can use them deliberately to distort public opinion via propaganda, or - more likely - without being aware of the tention between the symbols and their substantive message. Ultimately, by preserving citizens' trust in them and escaping from public blame, organizations can nurture their long-term positive reputation, which provides them political power and attenuates elected politicians' incentives to monitor and oversight them [@busuioc_2016; @bertelli_2018]. Hence, government communications with citizens can be a tool for enhancing democratic responsiveness and accountability, yet they can also use it as a means to undermine them.

@@

In addition, ELM has important methodological implications regarding the external validity of survey experiment studies of citizens' responses to government communications. The participants of these studies, including this particular one, have relatively high motivation to elaborate on the information that scholars presented them, because they may assume that they will be asked about it afterwards. In case they receive payment or other compensation for their participation, this further increases their motivation. By comparison, in the "real world", citizens are far less motivated to invest in thinking about government messages, and they probably also have lower ability to do so, given the high competition on their time and attention. Accordingly, their responses to performance information will less affected by asymetrical processing of information, because most of them are not likely to process this information in the first place. Instead, most of them will rely on simple cues and heuristic shortcuts. One can postulate that these differences exacerbate biases due to partisan and political ideology [@james_2017a; @baekgaard_2016] and implicit attitudes about the public-sector [@marvel_2015; @marvel_2015a; @hvidman_2016], while attenuating the role of negativity bias [@nielsen_2016; @olsen_2017]. Future studies may empirically explore these interactions. Regardless, scholars should also take to account this methodological issue when designing survey experiments, and attempt to simulate real world conditions, as much as possible. For instance, they can add filler sections in the survey before the real manipulation to attenuate participants' motivation and ability to elaborate. In addition, because the graphic design of communications also affects motivation and ability to elborate, as shown here, scholars should simulate the design of real communications.

Closing...

>>>>

To be presented at the 2018 Transatlantic Dialogue (TAD 14) at Workshop 1: *Behavioral Public Administration and Psychological perspectives*.

Preliminary draft (last edited at `r Sys.Date()`)[^footnote_scripts] - do not cite without author's permission.

[^footnote_scripts]: Original R scripts available at:  https://github.com/saaralonbarkat/symbols_and_information

@@

I conducted the survey experiment through Qualtrics Survey Software. To select and recruit residents of these specific cities, I was assisted by an Israeli internet research panel company *iPanel*. The online survey link was sent by iPanel, between 28.1-1.2.2018. A total of `r N_0` respondents completed the survey (a response rate of `r round((N_0/5077)*100,1)`%).[^footnote_responserate] Of these, I later filtered out `r N_0-N_1` observations (`r round(((N_0-N_1)/N_0)*100,1)`%) due to multiple entries from the same IP address (n=`r filter_IP`), failure at the instructional manipulation test (n=`r filter_IMC`), submitting the survey in less than 3 or more than 30 minutes (n=`r filter_time`), and respondents under the age of 18 (n=`r filter_age`). These filters are not significantly associated with the areas or the experimental conditions. Ultimately, following this rigorous screening, I ended up with a sizable sample of `r N_1` subjects, of whom `r SVIVA2_01_haifa%>%nrow()` live in the Haifa-bay cities, and `r SVIVA2_01_center%>%nrow()` live in the center cities.

@@

<br>
**Figure 3: Experimental design for policy plan communications**
![](D:/UNIVERSITY/R/SVIVA/experiment 2 01-2018/myfigures/experiment_flow_long.png)

@@

A growing body of Public-Administration research seeks to assess whether and how government organizations can increase citizens' trust through direct communications with citizens. This study contributes to this discussion, by pointing to the influence of symbolic elements in public communications, and unpacking the theoretical mechanisms underlying their effect. Using a rigorous experimental design, embedded in a realistic setting focused on the case of the Israeli Environmental Protection Ministry, I tested the effect of symbolic elements on trust, and its interactions with the substantive content of communications and with their perceived personal relevance. To the best of my knowledge, this is the first study to explore these interactions in the context of government communications.

I found that participants' exposure to the familiar symbols of EPM increased their trust in its policy plans, and that the effect was greater when they were presented with a weak policy plan. The latter interaction is partly explained by the negative effect of the symbols on the elaboration of the policy plans. These findings are based on a large sample of more than `r (SVIVA2_01%>%nrow() %>% round(-2))-100` participants, and are robust for various model specifications. These results, if generalizable, suggest that incorporating familiar symbolic elements in public communications increase citizens' trust in government organizations and their actions, via two mechanisms: First, they evoke general positive feelings that are transferred to the specific organizations and policies. Second, by putting citizens in a positive mood, they distract them from thinking about the substantive content of the communication, and thereby, they cause them to disregard its weaknesses. These two supplementary mechanisms are theorized using the framework of the social psychology elaboration likelihood model [@petty_1986].

I further postulated, based on ELM, that the effect of symbols is attenuated the more the communication is personally relevant for citizens' lives. However, this latter expectation was not supported by the results. Residents of the polluted Haifa-Bay, who perceive EPM policy plan for reducing air-pollution in that area as highly relevant for their lives, were affected by the symbols not less than those who reside elsewhere. If anything, the results suggest that the symbols had a greater effect on Haifa-bay subjects. How can we reconcile these unexpected findings with the proposed theory? Given that the moderating role of personal relevance on peripheral cues was substantiated by numerous studies in social psychology, I would not argue that we should abandon this postulation based on this one study. One plausible explanation is that although Haifa-bay residents perceived the air-pollution policy as more relevant, their motivation to elaborate was not high enough to cause them to avoid peripheral cues. Certainly, the air-pollution in their area concerns them a great deal, as they point out themselves. However this issue represents a *continuous* concern that they may have got used to over the time, as opposed a new issue that emerged recently. If this explanation holds, this entails that citizens may resist the emotional effect of symbols only in rare circumbstances, which are charecterized by very high levels of personal relevance. These latter findings are also consistent with the findings of @alonbarkat_2017. In that study, the scholars assessed citizens' trust in the government electricity company, and its shaping by its familiar symbol, given citizens' recent personal experience of prolonged blackouts. Whilst their study was not purposly designed to test the interaction between symbols and personal relevance, one can assume that those who personally experienced blackouts perceived electricity service as more personally relevant. Still, they found that citizens who experienced blackouts were influenced by the symbol, as much as those who did not.

Certainly, we must be cautious when drawing general conclusions from our findings, for several particular reasons. First, this study focuses, empirically, on one public organization, which invested greatly in symbolic communications, most prominently, the use of high-profile celebrity comedians in its public campaigns. Future studies may further examine the proposed theory by focusing on additional cases of public organizations and policy areas. Moreover, in the symbols treatment group, I used a bundle of symbols, without dissentangling their distinct effects. Future studies may compare the effectiveness of different types of symbolic elements. Finally, this study focuses on communication about policy plans, and therefore we must be cautious about generalizing from these findings to other types of communications. Specifically, future studies may seek to test my theory with regard to communications that include evidence about performance, and examine the interaction bewteen symbols and information on good/bad outcomes.

Keeping these limitations in mind, this study has several important implications for public administration research. Most importantly, my findings highlight the significat role of branding and symbolic communication in shaping citizens' attitudes about government. Citizens' responses to symbolic elements in communications thus far received relatively little attention in Public administration literature, compared with the abounded research on their responses to information. Yet, this study, and others, demonstrated that symbols can be fairly effective in increasing citizens' trust, and mitigating their negative attitudes [@alonbarkat_2017; @karens_2016; @teodoro_2018]. Furthermore, my findings suggest that symbols decrease citizens' "rational", systematic processing of substantive information. This is good news as far as we are concerned mainly about citizens' *undervaluation* of government performance, and the negative biases that undermine the effectiveness of positive information.[^footnote_biases_ELM]. On the other hand, the emotive effect of government symbols on public opinion is highly problematic, insofar as it challenges the democratic ideal that citizens form their opinions about their government and its actions on the basis of rational thinking [@dahl_1973; @soroka_2010]. Moreover, it becomes especially problematic when symbols are used by poorly performing organizations,  or organizations that lack professional expertiese. It those cases, symbolic elements can mitigate justified public criticism of government failures. Organizations can use them deliberately to distort public opinion via propaganda, or - more likely - without being aware of the tention between the symbols and their substantive message. Ultimately, by preserving citizens' trust in them and escaping from public blame, organizations can nurture their long-term positive reputation, which provides them political power and attenuates elected politicians' incentives to monitor and oversight them [@busuioc_2016; @bertelli_2018]. Hence, whereas government communication with citizens is normally perceived as a tool for enhancing democratic responsiveness and accountability, organizations can also use it as a means to undermine them.

@@

Finally, ELM has important methodological implications regarding the external validity of survey experiment studies of citizens' responses to government communications. The participants of these studies, including this particular one, have relatively high motivation to elaborate on the information that scholars presented them, because they may assume that they will be asked about it afterwards. In case they receive payment or other compensation for their participation, this further increases their motivation. By comparison, in the "real world", citizens are far less motivated to invest in thinking about government messages, and they probably also have lower ability to do so, given the high competition on their time and attention. Accordingly, their responses to performance information will less affected by asymetrical processing of information, because most of them are not likely to process this information in the first place. Instead, most of them will rely on simple cues and heuristic shortcuts. I would expect, tentatively, that these differences exacerbate biases due to partisan and political ideology [@james_2017a; @baekgaard_2016] and implicit attitudes about the public-sector [@marvel_2015; @marvel_2015a; @hvidman_2016], while attenuating the role of negativity bias [@nielsen_2016; @olsen_2017]. Future studies may empirically explore these interactions. Regardless, scholars should also take to account this methodological issue when designing survey experiments, and attempt to simulate real world conditions, as much as possible. For instance, they can add filler sections in the survey before the real manipulation to attenuate participants' motivation and ability to elaborate. In addition, because the graphic design of communications also affects motivation and ability to elborate, as shown here, scholars should simulate the design of real communications.
